One-Particle Density Matrix Polarisation Susceptibility Tensors

### 2018-08-12 papers/1_dmatpol/submit/paper-submit.tex
Major revision of the induced dipole moment origin-independence proof

The latter equality is true only if the basis set functions are orthogonal, i.e., 

@@

Such a condition is necessary in order to meet the charge conservation
requirement (see Appendix~\ref{a:orig-dep} for a detailed discussion).
Therefore orthogonality of AO's is assumed from that point, unless stated otherwise.

@@

Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to

@@

%    \right] } \;.

@@

%This can be recast to

@@

  \right\} \;,

@@

where the `$\Re$' symbol denotes the real part,

@@

 \mathbb{L}({\bf r}_Q) \equiv {\bf C}^{\dagger} \cdot \mathbb{M}({\bf r}_Q) \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \;.

@@

Note also the similarity between Eq.~\eqref{e:L-projector} and Eq.~\eqref{e:dmu-exact}.
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent of the origin. 
This guarantees that the theory fulfils the total charge conservation requirement.
Therefore, in later course of the paper the dipole origin will be omitted.

@@

Let $\left\{ {\bf F}^{(1)}({\bf r}), {\bf F}^{(2)}({\bf r}), \ldots, {\bf F}^{(N)}({\bf r}), \ldots \right\}$ 
be a set of $N_{\rm max}$ distinct and randomly sampled 
spatial distributions of electric field. It is assumed that
the exact difference one\hyp{}particle density matrices (with respect to the unperturbed state)
defined as
%
\begin{equation}
 \delta \overline{\bf D}^{(N)} \equiv \overline{\bf D}^{(N)} - \overline{\bf D}^{(0)}
\end{equation}
%
are known for each sample (overline symbolises the exact estimate).
Now, for each pair of the AO indices we construct the following 
parameterisation of Eq.~\eqref{e:final-model.General}:
%
\begin{multline}\label{e:final-model.General.Parameters}
 \delta D^{(N)} = \sum_{i }^M \Big\{ 
                              \sum_u^{x,y,z} S^{[1]}_{iu} F_{iu}^{(N)} \\
%               + \sum_{ij}^M \sum_{uw}^{x,y,z} \left( S^{[2]}_{iu} + S^{[2]}_{jw} \right) F_{iu}^{(N)} F_{jw}^{(N)}
                +             \sum_{u}^{x,y,z} \sum_{w<u} r_{uw} S^{[2]}_{iuw} F_{iu}^{(N)} F_{iw}^{(N)} 
                        + \ldots \Big\}
\end{multline}
%
(the Greek subscripts were omitted here for notational simplicity).
In the above equation, $B_u^{(i;1)} = S^{[1]}_{iu}$ and $B_{uw}^{(i;2)} = r_{uw} S^{[2]}_{iuw}$,
where $r_{uw}$ is the symmetry factor equal to 1 for diagonal elements and 2 for off\hyp{}diagonal
elements of $B_{uw}^{(i;2)}$.
The multiple parameter blocks
(${\bf S}^{[1]}$, ${\bf S}^{[2]}$ and so on)
appear in the first power, allowing for linear least\hyp{}squares regression.
The square bracket superscripts denote the block of the parameter space.

To determine the optimum set,
$
 {\bf S} = 
\begin{pmatrix}
{\bf S}^{[1]} &
{\bf S}^{[2]} & \cdots
\end{pmatrix}^T
$, we define a loss function $Z$ 
that is subject to the least\hyp{}squares minimisation,
%
\begin{equation}\label{e:Z}
 Z({\bf S}) = \sum_N^{N_{\rm max}} \left( \delta D^{(N)} - \delta \overline{D}^{(N)} \right)^2 \;.
\end{equation}
%
The Hessian of $Z$ computed with respect to the parameters 
is parameter\hyp{}independent (constant) and generally non\hyp{}singular
as long as the electric fields on all distributed sites are different.
Therefore, the exact solution for the optimal parameters is given by the Newton equation
%
\begin{equation}\label{e:Newton}
 {\bf S} = -{\bf H}^{-1} \cdot {\bf g} \;,
\end{equation}
%
where ${\bf g}$ and ${\bf H}$ are the gradient vector and the Hessian matrix, respectively.
The explicit forms of the gradient and Hessian up to second\hyp{}order 
are given in the Appendix~\ref{a:blocks}.
Note that in this case the dimensions of parameter space for the block 1 and 2 are
equal to $3M$ and $6M$, respectively.

@@

The electric field\hyp{}induced dipole and primitive quadrupole moments were evaluated from the following expressions:
%
\begin{subequations}\label{e:dmult}
  \begin{align}
   \delta \upmu_{u}   &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{u}{\beta}  \;,\\
   \delta \Theta_{uw} &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{(u-u_{\rm CM})(w-w_{\rm CM})}{\beta} \;, 
  \end{align}
\end{subequations}
%
where the subscript `CM' denotes the centre of mass of water molecule.
When the distributed polarisabilities were used instead of the polarisation\hyp{}induced 
difference density matrices, primitive induced quadrupole moments were computed from
the induced dipole moments (given in Eq.~\eqref{e:dmu-dpol}) according to
%
\begin{equation}
  \delta {\BM \Theta} = \sum_i^{\rm occ} 
   \left\{ 
       \delta {\BM \upmu}_i \otimes \left( {\bf r}_i - {\bf r}_{\rm CM} \right) 
      + \left( {\bf r}_i - {\bf r}_{\rm CM} \right) \otimes \delta {\BM \upmu}_i
   \right\} \;.
\end{equation}
%
Primitive quadrupole moments were subsequently transformed to their traceless forms by
%
\begin{equation}
  \delta \widetilde{\Theta}_{uw} = \delta \Theta_{uw} - \frac{1}{3} \delta_{uw} \Tr{\BM\Theta} \;.
\end{equation}
%
Methodology for the calculation of the electric field due to point charges
can be found elsewhere.\cite{Blasiak.Cho.JCP.2014}

@@

\begin{figure*}[h]
\includegraphics[width=\textwidth]{fig-2.eps}
\caption{\label{f:fig-2} {\bf Performance of the distributed DMS tensors of ground\hyp{}state water molecule in strong
non\hyp{}uniform electric field.}
In this Figure, `Ab Initio DMS' (panels A, D and G) 
refers to the model from Eq.~\eqref{e:final-model.HF} and
`Generalised DMS' (panels B, E and H) 
refers to the generalised model with $R_{\rm max}=2$ 
where $R_{\rm max}$ is the maximum power
in the expansion from Eq.~\eqref{e:final-model.General}.
%Calculations of the electric field\hyp{}induced interaction energies (A-C), dipole moments (D-F) and
%quadrupole moments (G-I). 
Comparison with the distributed dipole\hyp{}dipole polarisability model 
is also shown in this Figure in panels C, F and I. 
The ensemble of electric fields corresponds to the `S' set
from Tables~\ref{t:results} and \ref{t:fields}.}
\end{figure*}

@@

%In this work, the \emph{ab initio} model of the 
%one\hyp{}particle density matrix susceptibility tensors
%at the HF level was derived for a closed shell molecule 
%in a spatial electric field distribution. It was found 
%that the change in the density matrix element is approximately 
%linearly proportional to the electric field evaluated at the 
%distributed site associated with the centre of charge of the 
%localised molecular orbital. Subsequently, the \emph{ab initio} 
%model served as a basis for the formulation of the generalised 
%model that is, in principle, not limited to the HF theory,
%and takes into account also quadratic and higher\hyp{}order terms 
%with respect to the electric field. The generalised model, which 
%consisted of a set of adjustable parameters, defined the first\hyp{} 
%and second\hyp{}order density matrix susceptibilities that were 
%determined through the least\hyp{}squares linear regression 
%with respect to the training ensemble of electrostatically 
%perturbed states.
%
%The above theoretically developed models were then used 
%to calculate the electric field\hyp{}induced interaction 
%energies, dipole moments and quadrupole moments of water 
%molecule that is subjected to perturbation due to uniform 
%electric field, as well as various point charge environments 
%generating non\hyp{}uniform electric field of weak, moderate
%and strong magnitudes. It was found that, while the \emph{ab initio} model yields qualitatively incorrect
%interaction energies, the atom\hyp{}distributed generalised model can reach quantitative level of accuracy
%for prediction of interaction energies, dipole moments and even quadrupole moments
%once, at least quadratic terms with respect to the electric field, are taken into account. 
In this work, it was demonstrated that the generalised DMS models from Eq.~\eqref{e:final-model.General}
can accurately describe the polarization\hyp{}induced deformation of the one\hyp{}particle ground\hyp{}state electronic density
due to the uniform and non\hyp{}uniform distribution of electric field.
It was also shown that the accuracy of the model can be improved by extending the parameter space of the susceptibility tensors.
Even at the current state of development, the generalised quadratic DMS model
is of comparable energetic accuracy as the distributed polarisability model (RMSE of 0.4~kcal/mol in strong 
electric fields and less than 0.06~kcal/mol in moderate and weak fields).
%It is emphasised here that the model can be easily extended by taking into account
%third\hyp{}order terms as well if fine accuracy is required.
%Moreover, the generalized DMS model can be used to any type of one\hyp{}particle density matrices at arbitrary level of theory.
%This however falls out of scope of this contribution and it will be studied
%elsewhere. 
Moreover, any type of one\hyp{}particle density matrix could be treated by using
generalised DMS models, which is much more difficult to realise when a conventional multipole
model is used.
Therefore, it is believed that
the generalised DMS approach
has an important advantage over the multipole\hyp{}based approach, i.e., it provides more detailed
tensor description of the spatial polarisation of the electron density distribution, that could be
combined for example with density\hyp{}based methods of interaction energy calculations.\cite{Mandado.Hermida-Ramon.JCTC.2011}

@@

Multipole one\hyp{}electron integrals depend on the origin with respect to which they are evaluated.
However, the induced dipole moment defined in Eq.~\eqref{e:dmu-4-exact.linear-approximation} 
(or equivalently in Eq.~\eqref{e:dmu-l-vector})
has to be origin\hyp{}independent. Note that
%
\begin{equation}
 \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf r}_Q) 
 = \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf 0}) - \left[ {\bf C}^\dagger \right]_{i\alpha} {\bf r}_Q  
\end{equation}
%
and
%
\begin{equation}
 \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf r}_Q) 
 = \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf 0}) - \left[ {\bf C}^\dagger \cdot {\bf D} \right]_{i\alpha} {\bf r}_Q \;.
\end{equation}
%
But ${\bf C}^\dagger \cdot {\bf D}={\bf C}^\dagger \cdot {\bf C} \cdot {\bf C}^\dagger={\bf C}^\dagger$ in orthogonal
AO basis which implies that
%
\begin{multline}
   \left[ \widetilde{\mathbb{L}} \right]_{i\alpha} ({\bf r}_Q) 
 = \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf r}_Q) 
 - \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf r}_Q) \\
 = \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf 0})   
 - \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf 0})
 = \left[ \widetilde{\mathbb{L}} \right]_{i\alpha} ({\bf 0}) \;.
\end{multline}
%
Therefore, it is proved that the polarisation\hyp{}induced distributed dipole moments 
defined in Eq.~\eqref{e:dmu-4-exact.linear-approximation} are origin\hyp{}independent.
Thus, one can compute one\hyp{}electron dipole integrals with respect to any origin 
and resulting the susceptibilities ${\bf B}_{\alpha\beta}^{(i;1)}$ 
from Eq.~\eqref{e:susceptibility-B} will be uniquely defined.

@@

The gradient vector ${\bf g}$ and the Hessian matrix ${\bf H}$ 
are built from blocks associated with a particular type of parameters, i.e.,
%
\begin{equation}\label{e:Newton.gH}
 {\bf g} = 
\begin{pmatrix}
{\bf g}^{[1]} \\ 
{\bf g}^{[2]} 
\end{pmatrix} ,\quad
 {\bf H} = 
\begin{pmatrix}
{\bf H}^{[11]} & {\bf H}^{[12]}  \\ 
{\bf H}^{[21]} & {\bf H}^{[22]}  
\end{pmatrix} \;,
\end{equation}
%
where the block indices 1 and 2 correspond to the first\hyp{} and second\hyp{}order susceptibilities, respectively.
Note that the second derivatives of $\delta D^{(N)}$ 
with respect to the adjustable parameters vanish
due to the linear functional form of Eq.~\eqref{e:final-model.General.Parameters}.
Thus, the gradient element of the $r$th block and Hessian element of the $(rs)$th block read
%
\begin{subequations}
 \begin{align}
  g^{[r ]}    &\equiv \frac{\partial   Z}{\partial S^{[r]}} 
     =-2\sum_N \overline{\delta D}^{(N)}
               \frac{\partial   \left[ \delta D^{(N)} \right]}{\partial S^{[r]}} \;,\\
  H^{[rs]} &\equiv \frac{\partial^2 Z}{\partial S^{[r]} \partial S^{[s]}}  
     = 2\sum_N 
        \frac{\partial   \left[ \delta D^{(N)} \right]}{\partial S^{[r]}}
        \frac{\partial   \left[ \delta D^{(N)} \right]}{\partial S^{[s]}} \;.
 \end{align}
\end{subequations}
%
The explicit formulae for the gradient are
%
\begin{subequations}
 \begin{align}
  g^{[1]}_{ku} &=-2\sum_N \overline{\delta D}^{(N)} F^{(N)}_{ku} \;,\\
  g^{[2]}_{kuw} &=-2r_{uw} \sum_N \overline{\delta D}^{(N)} F^{(N)}_{ku} F^{(N)}_{kw} \;.
 \end{align}
\end{subequations}
%
The Hessian subsequently follows to be
%
\begin{subequations}
 \begin{align}
  H^{[11]}_{ku,lw} &= 2\sum_N F^{(N)}_{ku} F^{(N)}_{lw} \;,\\
  H^{[12]}_{ku,lu'w'} &= 2r_{u'w'} \sum_N F^{(N)}_{ku} F^{(N)}_{lu'} F^{(N)}_{lw'}  \;,\\
  H^{[22]}_{kuw,lu'w'} &= 2r_{uw} r_{u'w'} \sum_N F^{(N)}_{ku} F^{(N)}_{kw} F^{(N)}_{lu'} F^{(N)}_{lw'} \;.
 \end{align}
\end{subequations}
%
Note that due to the symmetry of the Hessian matrix, the block $21$
is a transpose of the block $12$. 
The composite indices $ku$ and $kuw$ are constructed from the distributed site index $k$
and the appropriate symmetry\hyp{}adapted ($w<u$) Cartesian component 
of a particular DMS tensor: $u$ for the first\hyp{}order,
and $uw$ for the second\hyp{}order susceptibility tensor, respectively.

>>>>

The latter equality is true in a special case when the basis set functions are orthogonal, i.e., 

@@

%Such a condition is necessary in order to meet the charge conservation
%requirement (see Appendix~\ref{a:orig-dep} for a detailed discussion).
%Therefore orthogonality of AO's is assumed from that point, unless stated otherwise.

@@

By substituting the McWeeny's result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} one can show that

@@

%    \right] } \;,

@@

%which can be recast to

@@

  \right\} \;.

@@

In the above expression, the `$\Re$' symbol denotes the real part,

@@

 \mathbb{L}({\bf r}_Q) \equiv {\bf C}^{(0)\dagger} \cdot \mathbb{M}({\bf r}_Q) \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \;.

@@

%Note also the similarity between Eq.~\eqref{e:L-projector} and Eq.~\eqref{e:dmu-exact}.

For the sake of simplicity, 
the orthogonality of AO's is assumed in this work, unless stated otherwise.
It is proven in the Appendix~\ref{a:orig-dep} that under such circumstances the
induced dipole moment obtained in Eq.~\eqref{e:L-projector} is independent of the origin. 
This guarantees that the theory fulfils the total charge conservation requirement and,
therefore, in later course of the paper, the dipole origin will be omitted.

@@

Let $\left\{ {\bf F}^{(1)}({\bf r}), {\bf F}^{(2)}({\bf r}), \ldots, {\bf F}^{(N)}({\bf r}), \ldots \right\}$ 
be a set of $N_{\rm max}$ distinct and randomly sampled 
spatial distributions of electric field. It is assumed that
the exact difference one\hyp{}particle density matrices (with respect to the unperturbed state)
defined as
%
\begin{equation}
 \delta \overline{\bf D}^{(N)} \equiv \overline{\bf D}^{(N)} - \overline{\bf D}^{(0)}
\end{equation}
%
are known for each sample (overline symbolises the exact estimate).
Now, for each pair of the AO indices we construct the following 
parameterisation of Eq.~\eqref{e:final-model.General}:
%
\begin{multline}\label{e:final-model.General.Parameters}
 \delta D^{(N)} = \sum_{i }^M \Big\{ 
                              \sum_u^{x,y,z} s^{[1]}_{iu} F_{iu}^{(N)} \\
%               + \sum_{ij}^M \sum_{uw}^{x,y,z} \left( s^{[2]}_{iu} + s^{[2]}_{jw} \right) F_{iu}^{(N)} F_{jw}^{(N)}
                +             \sum_{u}^{x,y,z} \sum_{w<u} r_{uw} s^{[2]}_{iuw} F_{iu}^{(N)} F_{iw}^{(N)} 
                        + \ldots \Big\}
\end{multline}
%
(the Greek subscripts were omitted here for notational simplicity).
In the above equation, $B_u^{(i;1)} = s^{[1]}_{iu}$ and $B_{uw}^{(i;2)} = r_{uw} s^{[2]}_{iuw}$,
where $r_{uw}$ is the symmetry factor equal to 1 for diagonal elements and 2 for off\hyp{}diagonal
elements of $B_{uw}^{(i;2)}$.
The multiple parameter blocks
(${\bf s}^{[1]}$, ${\bf s}^{[2]}$ and so on)
appear in the first power, allowing for linear least\hyp{}squares regression.
The square bracket superscripts denote the block of the parameter space.

To determine the optimum set,
$
 {\bf s} = 
\begin{pmatrix}
{\bf s}^{[1]} &
{\bf s}^{[2]} & \cdots
\end{pmatrix}^{\rm T}
$, we define a loss function $Z$ 
that is subject to the least\hyp{}squares minimisation,
%
\begin{equation}\label{e:Z}
 Z({\bf s}) = \sum_N^{N_{\rm max}} \left( \delta D^{(N)} - \delta \overline{D}^{(N)} \right)^2 \;.
\end{equation}
%
The Hessian of $Z$ computed with respect to the parameters 
is parameter\hyp{}independent (constant) and generally non\hyp{}singular
as long as the electric fields on all distributed sites are different.
Therefore, the exact solution for the optimal parameters is given by the Newton equation
%
\begin{equation}\label{e:Newton}
 {\bf s} = -{\bf H}^{-1} \cdot {\bf g} \;,
\end{equation}
%
where ${\bf g}$ and ${\bf H}$ are the gradient vector and the Hessian matrix, respectively.
The explicit forms of the gradient and Hessian up to second\hyp{}order 
are given in the Appendix~\ref{a:blocks}.
Note that in this case the dimensions of parameter space for the block 1 and 2 are
equal to $3M$ and $6M$, respectively.

@@

The electric field\hyp{}induced dipole and primitive quadrupole moments were evaluated from the following expressions:
%
\begin{subequations}\label{e:dmult}
  \begin{align}
   \delta \upmu_{u}   &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{u}{\beta}  \;,\\
   \delta \Theta_{uw} &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{(u-u_{\rm CM})(w-w_{\rm CM})}{\beta} \;, 
  \end{align}
\end{subequations}
%
where the subscript `CM' denotes the centre of mass of water molecule.
When the distributed polarisabilities were used instead of the polarisation\hyp{}induced 
difference density matrices, primitive induced quadrupole moments were computed from
the induced dipole moments (given in Eq.~\eqref{e:dmu-dpol}) according to
%
\begin{equation}\label{e:dTheta-dpol}
  \delta {\BM \Theta} = \sum_i^{\rm occ} 
   \left\{ 
       \delta {\BM \upmu}_i \otimes \left( {\bf r}_i - {\bf r}_{\rm CM} \right) 
      + \left( {\bf r}_i - {\bf r}_{\rm CM} \right) \otimes \delta {\BM \upmu}_i
   \right\} \;.
\end{equation}
%
Primitive quadrupole moments were subsequently transformed to their traceless forms by
%
\begin{equation}
  \delta \widetilde{\Theta}_{uw} = \delta \Theta_{uw} - \frac{1}{3} \delta_{uw} \Tr{\BM\Theta} \;.
\end{equation}
%
Methodology for the calculation of the electric field due to point charges
can be found elsewhere.\cite{Blasiak.Cho.JCP.2014}

@@

\begin{figure*}[h]
\includegraphics[width=\textwidth]{fig-2.eps}
\caption{\label{f:fig-2} {\bf Performance of the distributed DMS tensors of ground\hyp{}state water molecule in strong
non\hyp{}uniform electric field.}
In this Figure, `Ab Initio DMS' (panels A, D and G) 
refers to the model from Eq.~\eqref{e:final-model.HF} and
`Generalised DMS' (panels B, E and H) 
refers to the generalised model with $R_{\rm max}=2$ 
where $R_{\rm max}$ is the maximum power
in the expansion from Eq.~\eqref{e:final-model.General}.
Comparison with the distributed dipole\hyp{}dipole polarisability model %(Eq.~\eqref{e:dE-dpol})
is also shown in this Figure in panels C, F and I. 
The ensemble of electric fields corresponds to the `S' set
from Tables~\ref{t:results} and \ref{t:fields}.}
\end{figure*}

@@

%In this work, the \emph{ab initio} model of the 
%one\hyp{}particle density matrix susceptibility tensors
%at the HF level was derived for a closed shell molecule 
%in a spatial electric field distribution. It was found 
%that the change in the density matrix element is approximately 
%linearly proportional to the electric field evaluated at the 
%distributed site associated with the centre of charge of the 
%localised molecular orbital. Subsequently, the \emph{ab initio} 
%model served as a basis for the formulation of the generalised 
%model that is, in principle, not limited to the HF theory,
%and takes into account also quadratic and higher\hyp{}order terms 
%with respect to the electric field. The generalised model, which 
%consisted of a set of adjustable parameters, defined the first\hyp{} 
%and second\hyp{}order density matrix susceptibilities that were 
%determined through the least\hyp{}squares linear regression 
%with respect to the training ensemble of electrostatically 
%perturbed states.
%
%The above theoretically developed models were then used 
%to calculate the electric field\hyp{}induced interaction 
%energies, dipole moments and quadrupole moments of water 
%molecule that is subjected to perturbation due to uniform 
%electric field, as well as various point charge environments 
%generating non\hyp{}uniform electric field of weak, moderate
%and strong magnitudes. It was found that, while the \emph{ab initio} model yields qualitatively incorrect
%interaction energies, the atom\hyp{}distributed generalised model can reach quantitative level of accuracy
%for prediction of interaction energies, dipole moments and even quadrupole moments
%once, at least quadratic terms with respect to the electric field, are taken into account. 
In this work, it was demonstrated that the generalised DMS models from Eq.~\eqref{e:final-model.General}
can accurately describe the polarization\hyp{}induced deformation of the one\hyp{}particle ground\hyp{}state electronic density
due to the uniform and non\hyp{}uniform distribution of electric field.
It was also shown that the accuracy of the model can be improved by extending the parameter space of the susceptibility tensors.
Even at the current state of development, the generalised quadratic DMS model
is of comparable and quantitative energetic accuracy as the distributed polarisability model (RMSE of 0.4~kcal/mol in strong 
electric fields and less than 0.06~kcal/mol in moderate and weak fields).
%It is emphasised here that the model can be easily extended by taking into account
%third\hyp{}order terms as well if fine accuracy is required.
%Moreover, the generalized DMS model can be used to any type of one\hyp{}particle density matrices at arbitrary level of theory.
%This however falls out of scope of this contribution and it will be studied
%elsewhere. 
Moreover, any type of one\hyp{}particle density matrix could be treated by using
generalised DMS models, which is much more difficult to realise when a conventional multipole
model is used.
Therefore, it is believed that
the generalised DMS approach
has an important advantage over the multipole\hyp{}based approach, i.e., it provides more detailed
tensor description of the spatial polarisation of the electron density distribution, that could be
combined for example with density\hyp{}based methods of interaction energy calculations.\cite{Mandado.Hermida-Ramon.JCTC.2011}

@@

Multipole one\hyp{}electron integrals depend on the origin with respect to which they are evaluated.
However, the induced dipole moment defined in Eq.~\eqref{e:L-projector} 
has to be origin\hyp{}independent. Denoting the overlap AO integrals 
by $S_{\alpha\beta} = \braket{\alpha}{\beta}$,
it follows that $\mathbb{M}_u({\bf r}_Q) = \mathbb{M}_u({\bf 0}) - u_Q {\bf S}$
for $u=x,y,z$. Thus,
%
\begin{equation}
 \mathbb{L}_u({\bf r}_Q) = \mathbb{L}_u({\bf 0})
                          - u_Q {\bf C}^{(0)\dagger} \cdot {\bf S} \cdot \left( {\bf 1} - {\bf D}^{(0)}\right) \;,
\end{equation}
%
where
%
\begin{equation}
 \mathbb{L}_u({\bf 0}) = {\bf C}^{(0)\dagger} \cdot \mathbb{M}_u({\bf 0}) \cdot \left( {\bf 1} - {\bf D}^{(0)}\right) \;.
\end{equation}
%
But
%
\begin{equation}
  {\bf C}^{(0)\dagger} \cdot {\bf S} \cdot \left( {\bf 1} - {\bf D}^{(0)}\right) 
= {\bf C}^{(0)\dagger} \cdot \left( {\bf S} - {\bf 1} \right)
\end{equation}
%
since ${\bf C}^{(0)\dagger} \cdot {\bf S} \cdot {\bf C}^{(0)} = {\bf 1}$. 
In the simplest case of ${\bf S} = {\bf 1}$ (i.e., the AO basis functions are orthogonal) 
$\mathbb{L}_u({\bf r}_Q) = \mathbb{L}_u({\bf 0})$ which guarantees charge conservation.
When a non\hyp{}orthogonal AO basis is used, 
the following general condition must be satisfied:
%
\begin{equation}
 \Re \left\{
  \Tr{
  \left[
  \delta {\bf C} \cdot {\bf C}^{(0)\dagger} \cdot \left( {\bf S} - {\bf 1} \right)  
 \right]}
 \right\} = 0
\end{equation}
%
Due to the fact that ${\BM \Delta}$ considered here is accurate only up to first\hyp{}order, 
the above expression is not valid. 
Therefore, the polarisation\hyp{}induced distributed dipole moments 
defined in Eq.~\eqref{e:L-projector} are exactly origin\hyp{}independent
as long as the AO basis is orthogonalised.
In such a case, one can compute one\hyp{}electron dipole integrals with respect to any origin 
and the resulting susceptibilities $\mathbb{B}^{(i;1)}$ 
from Eq.~\eqref{e:susceptibility-B} will be uniquely defined. They can be back\hyp{}transformed
to the original (non\hyp{}orthogonal) AO basis, 
${\bf X} \cdot \mathbb{B}^{(i;1)} \cdot {\bf X}$, where $\bf X$ is an orthogonaliser.
The numerical tests confirmed that $\Tr\left[\delta{\bf D}\cdot{\bf S}\right] \sim 10^{-16}$
with $\delta{\bf D}$ given by Eq.~\eqref{e:final-model.HF} and susceptibilities transformed
to the non\hyp{}orthogonal AO basis.

@@

The gradient vector ${\bf g}$ and the Hessian matrix ${\bf H}$ 
are built from blocks associated with a particular type of parameters, i.e.,
%
\begin{equation}\label{e:Newton.gH}
 {\bf g} = 
\begin{pmatrix}
{\bf g}^{[1]} \\ 
{\bf g}^{[2]} 
\end{pmatrix} ,\quad
 {\bf H} = 
\begin{pmatrix}
{\bf H}^{[11]} & {\bf H}^{[12]}  \\ 
{\bf H}^{[21]} & {\bf H}^{[22]}  
\end{pmatrix} \;,
\end{equation}
%
where the block indices 1 and 2 correspond to the first\hyp{} and second\hyp{}order susceptibilities, respectively.
Note that the second derivatives of $\delta D^{(N)}$ 
with respect to the adjustable parameters vanish
due to the linear functional form of Eq.~\eqref{e:final-model.General.Parameters}.
Thus, the gradient element of the $r$th block and Hessian element of the $(rs)$th block read
%
\begin{subequations}
 \begin{align}
  g^{[r ]}    &\equiv \frac{\partial   Z}{\partial s^{[r]}} 
     =-2\sum_N \overline{\delta D}^{(N)}
               \frac{\partial   \left[ \delta D^{(N)} \right]}{\partial s^{[r]}} \;,\\
  H^{[rs]} &\equiv \frac{\partial^2 Z}{\partial s^{[r]} \partial s^{[s]}}  
     = 2\sum_N 
        \frac{\partial   \left[ \delta D^{(N)} \right]}{\partial s^{[r]}}
        \frac{\partial   \left[ \delta D^{(N)} \right]}{\partial s^{[s]}} \;.
 \end{align}
\end{subequations}
%
The explicit formulae for the gradient are
%
\begin{subequations}
 \begin{align}
  g^{[1]}_{ku} &=-2\sum_N \overline{\delta D}^{(N)} F^{(N)}_{ku} \;,\\
  g^{[2]}_{kuw} &=-2r_{uw} \sum_N \overline{\delta D}^{(N)} F^{(N)}_{ku} F^{(N)}_{kw} \;.
 \end{align}
\end{subequations}
%
The Hessian subsequently follows to be
%
\begin{subequations}
 \begin{align}
  H^{[11]}_{ku,lw} &= 2\sum_N F^{(N)}_{ku} F^{(N)}_{lw} \;,\\
  H^{[12]}_{ku,lu'w'} &= 2r_{u'w'} \sum_N F^{(N)}_{ku} F^{(N)}_{lu'} F^{(N)}_{lw'}  \;,\\
  H^{[22]}_{kuw,lu'w'} &= 2r_{uw} r_{u'w'} \sum_N F^{(N)}_{ku} F^{(N)}_{kw} F^{(N)}_{lu'} F^{(N)}_{lw'} \;.
 \end{align}
\end{subequations}
%
Note that due to the symmetry of the Hessian matrix, the block $21$
is a transpose of the block $12$. The composite indices $ku$ and $kuw$ 
are constructed from the distributed site index $k$ and the appropriate 
symmetry\hyp{}adapted ($w<u$) Cartesian component of a particular DMS tensor: 
$u$ for the first\hyp{}order, and $uw$ for the second\hyp{}order susceptibility 
tensor, respectively. The method described above can be easily extended 
to third and higher orders.
